---
title: "Joining by the closest date"
author: "Laura DeCicco"
output: 
  rmarkdown::html_vignette:
    fig_height: 5
    fig_width: 7
vignette: >
  %\VignetteIndexEntry{Joining by the closest date}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
library(dplyr)

knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE,
                      fig.height=7,
                      fig.width=7)


```

Increasingly there are high frequency sensor data available for water quality data. There's a common need to join the sensor and discrete data by the closest time. This article will discuss how to do that with common tools, then we'll put those techniques together in a function.

## dataRetrieval

Let's look at site "01646500", and a nearby site with a real-time nitrate-plus-nitrite sensor.

```{r getData, cache=TRUE}
library(dataRetrieval)

site_uv <- "01646500"
site_qw <- "USGS-01646580"
pcode_uv <- "99133"
pcode_qw <- "00631"
uv_data <- readNWISuv(site_uv, pcode_uv)
start_sensor_date <- as.Date(min(uv_data$dateTime))
qw_data <- readWQPqw(site_qw, pcode_qw,
                     startDate = start_sensor_date)

```

First let's trim down the data sets so that they are easy to print in this document. 

The sensor data ("uv" data) has 2 columns of data that are important. The first task is to combine those columns. This is rather unique to this particular site and probably won't need to be done generally.

```{r trimUVdata}
library(dplyr)

uv_trim <- uv_data |> 
  select(uv_date = dateTime, 
         val1 = X_SUNA...Discontinued._99133_00000,
         val2 = X_SUNA_99133_00000) |> 
  mutate(val_uv = if_else(is.na(val1), val2, val1)) |> 
  select(-val1, -val2)

```

```{r showuvTrim, echo=FALSE}
knitr::kable(head(uv_trim))
```

Next we'll clean up the discrete water quality "qw" data to make it easy to follow in this tutorial. 

```{r trimQWdata}
qw_trim <- qw_data |> 
  filter(ActivityTypeCode == "Sample-Routine",
         !is.na(ActivityStartDateTime)) |> 
  arrange(ActivityStartDateTime) |> 
  select(qw_date = ActivityStartDateTime,
         val_qw = ResultMeasureValue,
         det_txt = ResultDetectionConditionText)
```

```{r showqwtrim, echo=FALSE}
knitr::kable(head(qw_trim))
```

Now we'll use the `data.table` package to do a join to the nearest date. The code to do that is here:

```{r}
library(data.table)
# coerce to data.table and append join columns to preserve the original columns 
setDT(qw_trim)[, join_date := qw_date]

setDT(uv_trim)[, join_date := uv_date]
# rolling join
closest_dt <- uv_trim[qw_trim, on = .(join_date), roll = "nearest"]
```

`closest_dt` is a `data.table` object. It similar to a data.frame, but not identical. We can convert it to a data.frame and then use `dplyr` commands. Note: the whole analysis can be done via `data.table`, but most examples in `EGRET` have used `dplyr`, which is why we bring it back to data.frame. `dplyr` also has a `join_by(closest())` option, but it is more complicated because you can only specify the closeness in either the forward or backwards direction (and we want either direction).

We first calculate "delta_time" - the difference in time between the uv and qw data. Then, if that difference is greater than 24 hours, we'll substitute `NA`.

```{r}
qw_closest <- data.frame(closest_dt) |> 
  mutate(delta_time = difftime(qw_date, uv_date, units="hours"),
         val_uv = if_else(abs(as.numeric(delta_time)) >= 24, NA, val_uv)) |> 
  select(-join_date)

```


```{r showqwClosest, echo=FALSE}
knitr::kable(head(qw_closest))
```


## Putting it together

A more realistic scenario might be that we want to create an `EGRET` Sample data frame, and get the real-time water quality and flow values joined by the closest date/time.

We'll re-pull the real-time data, and this time include discharge:

```{r getMoreData, cache=TRUE}
library(EGRET)

# Sensor discharge:
uv_flow_qw <- readNWISuv(site_uv, c(pcode_uv, "00060"), 
                         startDate = start_sensor_date)

#special cleanup needed at this site:
uv_flow_qw2 <- uv_flow_qw |> 
  rename(val1 = X_SUNA...Discontinued._99133_00000,
         val2 = X_SUNA_99133_00000,
         rmk1 = X_SUNA...Discontinued._99133_00000_cd,
         rmk2 = X_SUNA_99133_00000_cd) |> 
  mutate(qw_val_uv = if_else(is.na(val1), val2, val1),
         qw_rmk_uv = if_else(is.na(rmk1), rmk2, rmk1)) |> 
  select(-val1, -val2, -rmk1, -rmk2)

```

Here we create a function that does the `data.table` join, and then continues the steps necessary to obtain a `Sample` data.frame.

The inputs are:

|Argument  | Description                             |  
|----------|----------------------------------------:|
|qw_data |  Data frame with discrete water quality data | 
|uv_flow_qw| Data frame with real-time (high frequency) water quality and/or flow data   |
|hour_threshold |Number of hours that the dates need to be within to match up |
| join_by_qw | Name of the date/time column in the qw_data data frame to join by  |
| join_by_uv | Name of the date/time column in the uv_flow_qw data frame to join by |
| qw_val | Name of the water quality value column in the qw_data data frame  |
| qw_rmk | Name of the water quality remark column in the qw_data data frame |
| qw_det_val | Name of the  |
| qw_val_uv  | Name of the water quality value column in the uv_flow_qw data frame |
| qw_rmk_uv  | Name of the water quality remark column in the uv_flow_qw data frame|
| flow_val   | Name of the flow value column in the uv_flow_qw data frame|
| flow_rmk   |Name of the flow remark column in the uv_flow_qw data frame |

Default values are provided that match the output of `readWQPqw`.

```{r defFunc}

join_qw_uv <- function(qw_data, # data from readWQP
                       uv_flow_qw, # data from readNWISuv
                       hour_threshold = 24, # hours threshold for joining
                       join_by_qw = "ActivityStartDateTime", 
                       join_by_uv = "dateTime",
                       qw_val = "ResultMeasureValue",
                       qw_rmk = "ResultDetectionConditionText",
                       qw_det_val = "DetectionQuantitationLimitMeasure.MeasureValue",
                       qw_val_uv, # water quality value column in uv data
                       qw_rmk_uv, # water quality remark column in uv data
                       flow_val = "X_00060_00000", # uv flow parameter
                       flow_rmk = "X_00060_00000_cd"){ # uv flow parameter cd
  
  req_cols <- c(join_by_qw, qw_val, qw_rmk, qw_det_val)
  if(!all(req_cols %in% names(qw_data))){
    stop(paste('qw_data missing columns:', req_cols[!req_cols %in% names(qw_data)]))
  }
  
  req_cols_uv <- c(join_by_uv)
  if(!all(req_cols_uv %in% names(uv_flow_qw))){
    stop(paste('uv_data missing columns:', req_cols_uv[!req_cols_uv %in% names(uv_flow_qw)]))
  }
  
  setDT(qw_data)[, eval(parse(text = paste("join_date :=", join_by_qw)))]
  
  x <- qw_data[order(join_by_qw)]
  
  setDT(uv_flow_qw)[, eval(parse(text = paste("join_date :=", join_by_uv)))]
  
  # rolling join
  x <- uv_flow_qw[qw_data, on = .(join_date), roll = "nearest"]
  
  setnames(x, c(qw_val, join_by_uv, join_by_qw, qw_rmk, qw_det_val),
           c("val_qw","uv_date", "qw_date", "qw_rmk", "qw_det_val"))
  
  x_tib <- as_tibble(x)
  
  if(!is.na(flow_val) | flow_val != ""){
    x_tib$flow_uv <- x_tib[[flow_val]]
  }
  if(!is.na(flow_rmk) | flow_rmk != ""){
    x_tib$flow_rmk_uv <- x_tib[[flow_rmk]]
  }
  
  if(!is.na(qw_val_uv) | qw_val_uv != ""){
    x_tib$qw_val_uv <- x_tib[[qw_val_uv]]
  }
  if(!is.na(qw_rmk_uv) | qw_rmk_uv != ""){
    x_tib$qw_rmk_uv <- x_tib[[qw_rmk_uv]]
  }

  toMatch <- c("NON-DETECT", "NON DETECT", "NOT DETECTED",
             "DETECTED NOT QUANTIFIED", "BELOW QUANTIFICATION LIMIT")
  
  x_tib <- x_tib |> 
    mutate(delta_time = difftime(qw_date, uv_date, units = "hours"),
         qw_val_uv = if_else(abs(as.numeric(delta_time)) >= hour_threshold, 
                          NA, qw_val_uv),
         qualifier = if_else(grepl(paste(toMatch,collapse="|"),
                                   toupper(qw_rmk)),
                             "<", ""),
         value = if_else(qualifier == "<", qw_det_val, val_qw),
         date = as.Date(qw_date)) |> 
    select(any_of(c("uv_date", "qw_date", "delta_time", "date",
           "qw_val_uv", "qw_rmk_uv",
           "value", "qualifier", 
           "flow_uv", "flow_rmk_uv"))) |> 
    rename(dateTime = qw_date) 
  

  compressedData <- EGRET::compressData(x_tib[, c("date",
                                                "qualifier",
                                                "value")],
                                        verbose = FALSE)
  Sample <- EGRET::populateSampleColumns(compressedData)
  Sample <- Sample |>
    left_join(x_tib |>
                select(-qualifier) |> 
                rename(qw_dateTime = dateTime,
                       uv_dateTime = uv_date,
                       Date = date, 
                       ConcHigh = value),
              by = c("Date", "ConcHigh"))
  
  return(Sample)
  
}


```

Running the function:

```{r runFunction, message=FALSE}


Sample <- join_qw_uv(qw_data = qw_data,
                     uv_flow_qw =  uv_flow_qw2, 
                     hour_threshold = 24,
                     join_by_qw = "ActivityStartDateTime",
                     join_by_uv = "dateTime",
                     qw_val_uv = "qw_val_uv",
                     qw_rmk_uv = "qw_rmk_uv",
                     flow_val = "X_00060_00000",
                     flow_rmk = "X_00060_00000_cd")
```


In addition to the standard `Sample` output, you'll get:

|Column  | Description                             |  
|----------|----------------------------------------:|
| uv_dateTime | The date time from the uv data that was closest to the qw data. |
| qw_dateTime | The date time from the qw data frame. |
| delta_time | The difference in time in hours. |
| qw_val_uv | The water-quality uv value that was closest to the qw data. |
| qw_rmk_uv | The water-quality uv remark that was closest to the qw data. |
| flow_uv | The flow value closest to the qw data. |
| flow_rmk_uv | The flow remark closest to the qw data. |


Using the function without real-time flow data:

```{r noFlow, warning=FALSE}

Sample2 <- join_qw_uv(qw_data = qw_data,
                     uv_flow_qw =  uv_flow_qw2, 
                     hour_threshold = 24,
                     join_by_qw = "ActivityStartDateTime",
                     join_by_uv = "dateTime",
                     qw_val_uv = "qw_val_uv",
                     qw_rmk_uv = "qw_rmk_uv",
                     flow_val = "",
                     flow_rmk = "")

```

Using the function without real-time qw data:

```{r noQW, warning=FALSE}

Sample3 <- join_qw_uv(qw_data = qw_data,
                     uv_flow_qw =  uv_flow_qw2, 
                     hour_threshold = 24,
                     join_by_qw = "ActivityStartDateTime",
                     join_by_uv = "dateTime",
                     qw_val_uv = "",
                     qw_rmk_uv = "",
                     flow_val = "X_00060_00000",
                     flow_rmk = "X_00060_00000_cd")

```
