---
title: "Trends by month"
author: "Robert M. Hirsch"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Trends by month}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This vignette is about producing a single type of figure.  It is designed to examine the trends between a pair of years (using **runPairs** in EGRET terminology) to see what the size of the trend is for each of the twelve months of the year.  It was used to make Figure 8 in Choquette et al., 2019.  It is very helpful in trying to see if the trend is very specific to a certain part of the year (or even opposite in different parts of the year) versus being fairly similar across all the months.

## How to do it

You need two functions.  One is called **runPairsMonths.R** to do the computations and the other is called **plotOne.R** which makes the plots.  These two functions are included below.  You should set up a working directory and then you can copy these functions into your working directory.  You be you need an EGRET eList in the directory as well.  It should be an eList for which modelEstimation has already been run.

Here is the code

```{r include=TRUE, eval = TRUE}
# This is function called runPairsMonths
# It is virtually identical to the EGRET function runPairs
# To learn the meaning of each of these initial arguments
# The best thing to do is load the EGRET package
# and then do ?runPairs
# the help page you will get defines all of these
runPairsMonths <- function(eList, year1, year2, windowSide = 0, flowBreak = FALSE,
         Q1EndDate = NA, QStartDate = NA, QEndDate = NA, wall = FALSE,
         oldSurface = TRUE, sample1EndDate = NA, sampleStartDate = NA,
         sampleEndDate = NA, minNumObs = 100,
         minNumUncen = 50, fractMin = 0.75, windowY = 7, windowQ = 2,
         windowS = 0.5, edgeAdjust = TRUE){
paStart <- 1
paLong <- 12 

{
  if (wall & oldSurface) {
    message("Setting both arguments wall and oldSurfaces to TRUE are not allowed.")
    message("Re-calculating surface.")
    oldSurface <- FALSE
  }
  if (!is.egret(eList)) {
    stop("Please check eList argument")
  }
  localSample <- getSample(eList)
  startEndSurface1 <- startEnd(paStart, paLong, year1)
  startEndSurface2 <- startEnd(paStart, paLong, year2)
  if (startEndSurface2$startDate > range(localSample$Date)[2]) {
    stop("year2 is outside the Sample range")
  }
  if (startEndSurface1$endDate < range(localSample$Date)[1]) {
    stop("year1 is outside the Sample range")
  }
  sampleStartDate <- if (is.na(sampleStartDate)) 
    localSample$Date[1]
  else as.Date(sampleStartDate)
  numSamples <- length(localSample$Date)
  sampleEndDate <- if (is.na(sampleEndDate)) 
    localSample$Date[numSamples]
  else as.Date(sampleEndDate)
  localDaily <- getDaily(eList)
  QStartDate <- if (is.na(QStartDate)) 
    localDaily$Date[1]
  else as.Date(QStartDate)
  numQDays <- length(localDaily$Date)
  QEndDate <- if (is.na(QEndDate)) 
    localDaily$Date[numQDays]
  else as.Date(QEndDate)
  if (sampleStartDate > as.Date(startEndSurface1[[2]])) {
    stop("Sample start is later than year2")
  }
  if (sampleEndDate < as.Date(startEndSurface2[[1]])) {
    stop("Sample end is earlier than year1")
  }
  localsurfaces <- getSurfaces(eList)
  if (oldSurface) {
    if (all(is.na(localsurfaces))) {
      message("No surface included in eList, running estSurface function")
      oldSurface <- FALSE
    }
  }
  if (flowBreak && is.na(Q1EndDate)) 
    stop("if there is a flowBreak you must provide Q1EndDate")
  flowNormStartCol <- "flowNormStart"
  flowNormEndCol <- "flowNormEnd"
  flowStartCol <- "flowStart"
  flowEndCol <- "flowEnd"
  if (windowSide <= 0 && !flowBreak) {
    flowStart <- c(startEndSurface1[["startDate"]], startEndSurface2[["startDate"]])
    flowEnd <- c(startEndSurface1[["endDate"]], startEndSurface2[["endDate"]])
    flowNormStart <- c(QStartDate, QStartDate)
    flowNormEnd <- c(QEndDate, QEndDate)
    dateInfo <- data.frame(flowNormStart, flowNormEnd, flowStart, 
                           flowEnd, stringsAsFactors = FALSE)
  }
  else if (windowSide > 0 & !flowBreak) {
    dateInfo1 <- makeDateInfo(windowSide, startEndSurface1[["startDate"]], 
                              startEndSurface1[["endDate"]], QStartDate, QEndDate)
    dateInfo2 <- makeDateInfo(windowSide, startEndSurface2[["startDate"]], 
                              startEndSurface2[["endDate"]], QStartDate, QEndDate)
    dateInfo <- rbind(dateInfo1, dateInfo2)
  }
  else if (windowSide <= 0 && flowBreak) {
    Q1EndDate <- as.Date(Q1EndDate)
    Q2StartDate <- Q1EndDate + 1
    flowStart <- c(startEndSurface1[["startDate"]], startEndSurface2[["startDate"]])
    flowEnd <- c(startEndSurface1[["endDate"]], startEndSurface2[["endDate"]])
    flowNormStart <- c(as.Date(QStartDate), as.Date(Q2StartDate))
    flowNormEnd <- c(as.Date(Q1EndDate), as.Date(QEndDate))
    dateInfo <- data.frame(flowNormStart, flowNormEnd, flowStart, 
                           flowEnd, stringsAsFactors = FALSE)
  }
  else {
    Q1EndDate <- as.Date(Q1EndDate)
    Q2StartDate <- Q1EndDate + 1
    dateInfo1 <- makeDateInfo(windowSide, startEndSurface1[["startDate"]], 
                              startEndSurface1[["endDate"]], QStartDate, Q1EndDate)
    dateInfo2 <- makeDateInfo(windowSide, startEndSurface2[["startDate"]], 
                              startEndSurface2[["endDate"]], Q2StartDate, QEndDate)
    dateInfo <- rbind(dateInfo1, dateInfo2)
  }
  if (wall) {
    if (is.na(sample1EndDate)) {
      stop("When using the wall option, please specify sample1EndDate")
    }
    sample1EndDate <- as.Date(sample1EndDate)
    sample2StartDate <- as.Date(sample1EndDate) + 1
    sample1StartDate <- as.Date(sampleStartDate)
    sample2EndDate <- as.Date(sampleEndDate)
  }
  else {
    sample1StartDate <- as.Date(sampleStartDate)
    sample2StartDate <- as.Date(sampleStartDate)
    sample1EndDate <- as.Date(sampleEndDate)
    sample2EndDate <- as.Date(sampleEndDate)
  }
  Sample1 <- localSample[localSample$Date >= sample1StartDate & 
                           localSample$Date <= sample1EndDate, ]
  Sample2 <- localSample[localSample$Date >= sample2StartDate & 
                           localSample$Date <= sample2EndDate, ]
  fractMin <- min(fractMin, 1)
  minNumObs <- ceiling(min(minNumObs, fractMin * length(Sample1$Date), 
                           fractMin * length(Sample2$Date)))
  minNumUncen <- ceiling(min(0.5 * minNumObs, minNumUncen))
  message("Sample1 has ", length(Sample1$Date), " Samples and ", 
          sum(Sample1$Uncen), " are uncensored")
  message("Sample2 has ", length(Sample2$Date), " Samples and ", 
          sum(Sample2$Uncen), " are uncensored")
  message("minNumObs has been set to ", minNumObs, " minNumUncen has been set to ", 
          minNumUncen)
  check <- rep(1, 4)
  if (minNumObs > length(Sample1$Date)) 
    check[1] <- 0
  if (minNumObs > length(Sample2$Date)) 
    check[2] <- 0
  if (minNumUncen > sum(Sample1$Uncen)) 
    check[3] <- 0
  if (minNumUncen > sum(Sample2$Uncen)) 
    check[4] <- 0
  if (sum(check) < 4) {
    stop("Data set too small for minNumObs or minNumUncen")
  }
  message("Sample1 has ", length(Sample1$Date), " Samples and ", 
          sum(Sample1$Uncen), " are uncensored")
  message("Sample2 has ", length(Sample2$Date), " Samples and ", 
          sum(Sample2$Uncen), " are uncensored")
  message("minNumObs has been set to ", minNumObs, " minNumUncen has been set to ", 
          minNumUncen)
  Daily1 <- localDaily[localDaily$Date >= dateInfo$flowNormStart[1] & 
                         localDaily$Date <= dateInfo$flowNormEnd[1], ]
  Daily2 <- localDaily[localDaily$Date >= dateInfo$flowNormStart[2] & 
                         localDaily$Date <= dateInfo$flowNormEnd[2], ]
  if (oldSurface) {
    checkSurfaceSpan(eList)
    if (all(c("Year", "LogQ", "surfaceIndex") %in% names(attributes(localsurfaces)))) {
      surfaceYear <- attr(localsurfaces, "Year")
      LogQ <- attr(localsurfaces, "LogQ")
    }
    else {
      localINFO <- getInfo(eList)
      LogQ <- seq(localINFO$bottomLogQ, by = localINFO$stepLogQ, 
                  length.out = localINFO$nVectorLogQ)
      surfaceYear <- seq(localINFO$bottomYear, by = localINFO$stepYear, 
                         length.out = localINFO$nVectorYear)
    }
    startDec1 <- decimalDate(startEndSurface1[["startDate"]])
    endDec1 <- decimalDate(startEndSurface1[["endDate"]])
    startDec2 <- decimalDate(startEndSurface2[["startDate"]])
    endDec2 <- decimalDate(startEndSurface2[["endDate"]])
    surfIndex1 <- which(surfaceYear >= startDec1 & surfaceYear <= 
                          endDec1)
    surfIndex1 <- c(surfIndex1[1] - 1, surfIndex1, surfIndex1[length(surfIndex1)] + 
                      1)
    surfIndex2 <- which(surfaceYear >= startDec2 & surfaceYear <= 
                          endDec2)
    surfIndex2 <- c(surfIndex2[1] - 1, surfIndex2, surfIndex2[length(surfIndex2)] + 
                      1)
    surfaces1 <- localsurfaces[, surfIndex1, ]
    surfaces2 <- localsurfaces[, surfIndex2, ]
    attr(surfaces1, "LogQ") <- LogQ
    attr(surfaces1, "Year") <- surfaceYear[surfIndex1]
    attr(surfaces2, "LogQ") <- LogQ
    attr(surfaces2, "Year") <- surfaceYear[surfIndex2]
  }
  else {
    surfaces1 <- estSurfaces(eList, surfaceStart = startEndSurface1[["startDate"]], 
                             surfaceEnd = startEndSurface1[["endDate"]], localSample = Sample1, 
                             minNumObs = minNumObs, minNumUncen = minNumUncen, 
                             windowY = windowY, windowQ = windowQ, windowS = windowS, 
                             edgeAdjust = edgeAdjust, verbose = FALSE)
    surfaces2 <- estSurfaces(eList, surfaceStart = startEndSurface2[["startDate"]], 
                             surfaceEnd = startEndSurface2[["endDate"]], localSample = Sample2, 
                             minNumObs = minNumObs, minNumUncen = minNumUncen, 
                             windowY = windowY, windowQ = windowQ, windowS = windowS, 
                             edgeAdjust = edgeAdjust, verbose = FALSE)
  }
  DailyRS1FD1 <- estDailyFromSurfaces(eList, localsurfaces = surfaces1, 
                                      localDaily = Daily1)
  annualFlex <- setupYears(DailyRS1FD1, paLong = paLong, paStart = paStart)
  c11 <- mean(annualFlex$FNConc, na.rm = TRUE)
  f11 <- mean(annualFlex$FNFlux, na.rm = TRUE)
  DailyRS2FD2 <- estDailyFromSurfaces(eList, localsurfaces = surfaces2, 
                                      localDaily = Daily2)
  annualFlex <- setupYears(DailyRS2FD2, paLong = paLong, paStart = paStart)
  c22 <- mean(annualFlex$FNConc, na.rm = TRUE)
  f22 <- mean(annualFlex$FNFlux, na.rm = TRUE)
  Daily0 <- localDaily[localDaily$Date >= QStartDate & localDaily$Date <= 
                         QEndDate, ]
  DailyRS1FD0 <- estDailyFromSurfaces(eList, localsurfaces = surfaces1, 
                                      localDaily = Daily0)
  annualFlex <- setupYears(DailyRS1FD0, paLong = paLong, paStart = paStart)
  c10 <- mean(annualFlex$FNConc, na.rm = TRUE)
  f10 <- mean(annualFlex$FNFlux, na.rm = TRUE)
  DailyRS2FD0 <- estDailyFromSurfaces(eList, localsurfaces = surfaces2, 
                                      localDaily = Daily0)
  annualFlex <- setupYears(DailyRS2FD0, paLong = paLong, paStart = paStart)
  c20 <- mean(annualFlex$FNConc, na.rm = TRUE)
  f20 <- mean(annualFlex$FNFlux, na.rm = TRUE)
  cDeltaTotal <- c22 - c11
  cRSpart <- c20 - c10
  cFDpart <- cDeltaTotal - cRSpart
  fDeltaTotal <- f22 - f11
  fRSpart <- f20 - f10
  fFDpart <- fDeltaTotal - fRSpart
  totChangePct_conc <- 100 * cDeltaTotal/c11
  totChangePct_flux <- 100 * fDeltaTotal/f11
  CQTC_percent_conc <- 100 * cRSpart/c11
  QTC_percent_conc <- 100 * cFDpart/c11
  CQTC_percent_flux <- 100 * fRSpart/f11
  QTC_percent_flux <- 100 * fFDpart/f11
  pairResults <- as.data.frame(matrix(ncol = 7, nrow = 2))
  colnames(pairResults) <- c("TotalChange", "CQTC", "QTC", 
                             "x10", "x11", "x20", "x22")
  rownames(pairResults) <- c("Conc", "Flux")
  pairResults[1, ] <- c(cDeltaTotal, cRSpart, cFDpart, c10, 
                        c11, c20, c22)
  pairResults[2, ] <- 0.00036525 * c(fDeltaTotal, fRSpart, 
                                     fFDpart, f10, f11, f20, f22)
  yearPairInfo <- c(paStart, paLong, year1, year2)
  names(yearPairInfo) <- c("paStart", "paLong", "year1", "year2")
  attr(pairResults, "yearPair") <- yearPairInfo
  attr(pairResults, "dateInfo") <- dateInfo
  SampleBlocks <- c(sample1StartDate, sample1EndDate, sample2StartDate, 
                    sample2EndDate)
  names(SampleBlocks) <- c("sample1StartDate", "sample1EndDate", 
                           "sample2StartDate", "sample2EndDate")
  attr(pairResults, "SampleBlocks") <- SampleBlocks
  Other <- list(minNumObs = minNumObs, minNumUncen = minNumUncen, 
                windowY = windowY, windowQ = windowQ, windowS = windowS, 
                wall = wall, edgeAdjust = edgeAdjust, QStartDate = as.Date(QStartDate), 
                QEndDate = as.Date(QEndDate), PercentChangeConc = c(`Total Percent Change` = totChangePct_conc, 
                                                                    `CQTC Percent` = CQTC_percent_conc, `QTC Percent` = QTC_percent_conc), 
                PercentChangeFlux = c(`Total Percent Change` = totChangePct_flux, 
                                      `CQTC Percent` = CQTC_percent_flux, `QTC Percent` = QTC_percent_flux))
  attr(pairResults, "Other") <- Other
  cat("\n  ", eList$INFO$shortName, "\n  ", eList$INFO$paramShortName)
  periodName <- setSeasonLabelByUser(paStart, paLong)
  cat("\n  ", periodName, "\n")
  if (wall) 
    cat("\n Sample data set was partitioned with a wall right after ", 
        as.character(sample1EndDate), "\n")
  cat("\n Change estimates ", year2, " minus ", year1, "\n")
  totChange <- format(pairResults[1, 1], digits = 3)
  totChangePct_conc_f <- format(totChangePct_conc, digits = 2)
  cat("\n For concentration: total change is ", totChange, 
      "mg/L")
  cat("\n expressed as Percent Change is ", totChangePct_conc_f, 
      "%")
  pctRS <- format(CQTC_percent_conc, digits = 2)
  pctFD <- format(QTC_percent_conc, digits = 2)
  cat("\n\n Concentration v. Q Trend Component ", pctRS, "%\n       Q Trend Component            ", 
      pctFD, "% \n\n")
  totChange <- format(pairResults[2, 1], digits = 3)
  totChangePct_flux_f <- format((totChangePct_flux), digits = 2)
  cat("\n For flux: total change is ", totChange, "million kg/year")
  cat("\n expressed as Percent Change is ", totChangePct_flux_f, 
      "%")
  pctRS <- format(CQTC_percent_flux, digits = 2)
  pctFD <- format(QTC_percent_flux, digits = 2)
  cat("\n\n Concentration v. Q Trend Component ", pctRS, "%\n       Q Trend Component            ", 
      pctFD, "% \n\n")
  print(pairResults[, 1:7], digits = 2)
  z <- matrix(ncol = 12, nrow = 2)
  colnames(z) <- c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
  k <- c(31, 28.25, 31, 30, 31, 30, 31, 31, 30, 31, 30, 31) / eList$INFO$drainSqKm
  eList1 <- as.egret(eList$INFO, DailyRS1FD1, eList$Sample)
  monthlyResults1 <- calculateMonthlyResults(eList1)
  monthlyResults1 <- na.omit(monthlyResults1)
  eList2 <- as.egret(eList$INFO, DailyRS2FD2, eList$Sample)
  monthlyResults2 <- calculateMonthlyResults(eList2)
  monthlyResults2 <- na.omit(monthlyResults2)
  for(i in 1:12){
    m1 <- monthlyResults1[monthlyResults1[,1]==i,]
    z[1,i] <- m1[1,8] * k[i]
    m2 <- monthlyResults2[monthlyResults2[,1]==i,]
    z[2,i] <- m2[1,8] * k[i]  
  }
}
return(z)
}
#
#
# this is written for kg/month/km^2
# the arguments yMax, arrowFactor, and name can all be
# adjusted to make some changes in the figure
plotOne <- function(z, yMax = NA, arrowFactor = 0.4, 
                    name = "Change in yield, by month"){
zMax <- max(z)
yMax <- if(is.na(yMax)) zMax * 1.05 else yMax
monthAbb <- c("J","F","M","A","M","J","J","A","S","O","N","D")
par(las = 1, tck = 0.02, xaxs = "i", yaxs = "i")
plot(seq(1,12),z[1,],xlim = c(0.5,12.5), ylim = c(0, yMax),
      xlab = "", ylab = "Change in yield, kg / month / km^2",
     main = name, col = "black", axes = FALSE, cex.lab = 0.95)
axis(1, at = seq(1, 12), labels = monthAbb, tick = TRUE)
axis(2, at = NULL, labels = TRUE, tick = TRUE)
axis(3, at = seq(1, 12), labels = FALSE, tick = TRUE)
axis(4, at = NULL, labels = FALSE, tick = TRUE)
box()
par(new = TRUE)
plot(seq(1,12),z[2,],xlim = c(0.5,12.5), ylim = c(0, yMax),
     xlab = "", ylab = "",
     main = "", col = "red", axes = FALSE)
for(m in 1: 12){
  x0 <- m
  x1 <- m
  y0 <- z[1,m]
  y1 <- z[2,m]
  col <- if(y1 > y0) "red" else "black"
  length <- arrowFactor * abs(y1 - y0) 
  angle <- 30
  lwd <- 2
  arrows(x0, y0, x1, y1, length = length, angle = angle, col = col, lwd = lwd)
}
}
```

Now let's run it

```{r eval = TRUE, echo = TRUE}

library(EGRET)
load("Conowingo.PO4.RData")
# source("runPairsMonths.R")
# source("plotOne.R")
z <- runPairsMonths(eList, year1 = 2006, year2 = 2017)
# z contains the results, they are expressed as yields in kg/month/km^2
# z[1,1:12] are the yields in year 1, by month Jan, Feb, ... Dec
# z[2,1:12] are the yields in year 2, by month Jan, Feb, ... Dec
print(z)
#
plotOne(z)
####
###we might want to make a few changes in the look of the figure, for example

plotOne(z, yMax = 1.2, arrowFactor = 0.6, name = "Change in yield between 2006 and 2017\nPO4 at Conowingo")
```

# Reference cited

Choquette, A.F., Hirsch, R.M., Murphy, J.C., Johnson, L.T. and Confesor Jr, R.B., 2019. Tracking changes in nutrient delivery to western Lake Erie: Approaches to compensate for variability and trends in streamflow. Journal of Great Lakes Research, 45(1), pp.21-39.

