---
title: "Trends by month"
author: "Robert M. Hirsch"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trends by month}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.width = 7,
                      fig.height = 7)
```

## Introduction

This vignette is about producing a single type of figure.  It is designed to examine the trends between a pair of years (using **runPairs** in EGRET terminology) to see what the size of the trend is for each of the twelve months of the year.  It was used to make Figure 8 in Choquette et al., 2019.  It is very helpful in trying to see if the trend is very specific to a certain part of the year (or even opposite in different parts of the year) versus being fairly similar across all the months.

As of `EGRET` version 3.0.7.2 (or CRAN version 3.0.8), the output of `runPairs` includes information about the trends per individual month. This output is attached via an attribute called "byMonth".



```{r eval = TRUE, echo = TRUE, message=FALSE, fig.height=4}

library(EGRET)
load("Conowingo.PO4.RData")
# source("runPairsMonths.R")
# source("plotOne.R")
pairResults <- runPairs(eList, windowSide = 0,
                        paStart = 1, paLong = 12,
                        year1 = 2006, year2 = 2017)
```



To see the tabular output of the results, use the "byMonthly" attribute from the `runPairs` result. Here we show the a few of the months output:

```{r}
monthly <-  attr(pairResults, "byMonth")

knitr::kable(monthly, digits = 3)

```

Next, let's plot the results using the custom function from above:

```{r}
plotMonthTrend(pairResults)

```

We might want to make a few changes in the look of the figure, for example:

```{r eval = TRUE, echo = TRUE, fig.height=4}
plotMonthTrend(pairResults,
               yMax = 1.2, 
               arrowFactor = 0.6)
```

We could also look at concentration:

```{r eval = TRUE, echo = TRUE, fig.height=4}
plotMonthTrend(pairResults, 
               arrowFactor = 1.5,
               flux = FALSE)
```

# Reference cited

Choquette, A.F., Hirsch, R.M., Murphy, J.C., Johnson, L.T. and Confesor Jr, R.B., 2019. Tracking changes in nutrient delivery to western Lake Erie: Approaches to compensate for variability and trends in streamflow. Journal of Great Lakes Research, 45(1), pp.21-39.

